---
title: "ocEAn Intro"
author: "A. Dugourd"
date: "12/01/2023"
output:
  html_document:
    toc: true
    toc_depth: 2
    keep_md: true
vignette: >
  %\VignetteIndexEntry{ocEAn Intro}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

OCEAN: a method that defines metabolic enzyme footprint from a curated reduced
version of the recon2 reaction network. The metabolic enzyme footprints
are used to explore coordinated deregulations of metabolite abundances
with respect to their position relative to metabolic enzymes.
This is similar to Kinase-substrate and TF-targets enrichment analyses.

## install ocEAn

```
library(devtools)
install_github("saezlab/ocean")
```


## Tutorial with a kidney cancer toy metabolomic dataset

```
library(ocean)
```

## Differential analysis

With data from: https://www.embopress.org/doi/full/10.15252/msb.20209730.

```
unique(toy_targets$condition)
comparisons <- list('tumorVsHealthy' = c(1,-2))

limmaRes <- runLimma(measurements = toy_metabolomic_data,
                     targets = toy_targets,
                     comparisons = comparisons)
```

This differential anaylsis result represents metabolic up and down-regulations
of metabolite abundances in kidney tumor tissue compared to adjacent
non-tumoral tissue.

## Format differential analysis result

In order to assess the metabolic imbalance, we use the t statistic of the
differential analysis. The t-statistic represents a good estimate of the
relative significance of up and down-regulations of metabolic abundances.

```
t_table <- ttop_list_to_t_table(
  limma_res_to_ttop_list(limma_res = limmaRes,
                         comp_names = names(comparisons),
                         number = length(toy_metabolomic_data[,1]),
                         adjust.method = "fdr"))
```

## Next, the metabolic names have to be converted to a standard identifer.

In this case, we use KEGG IDs.
This step is particularly important because this is where the users
metabolic identifiers are mapped to the kegg ids used by the method.

**THE USER SHOULD PROVIDE A MAPPING TABLE IN THE SAME FORMAT AS THE ONE
PRESENTED HERE.** (You can look at it to inspire yourself from it.) The
mapping table should map the users own metabolic identifiers to kegg compound
IDs.

```
t_table <- t_table_metactivity_input_formater(
  metabolomic_t_table = t_table,
  mapping_table = mapping_table,
  affixes = c("c","l","x","m","e","n","r")
)
```

## These are the available pathways to choose from

```
View(unique(recon2_redhuman$pathway))
```

## Subnetworks

```
data(expressed_genes)
```

In this case, we will use all pathways available. Users can restrict this
list on a per-need basis.

```
all_pathways <- unique(recon2_redhuman$pathway)
sub_network <- model_to_pathway_sif(pathway_to_keep = all_pathways$X1)
```

Some formating of the network is required, especially to make the method
easier to run and the results easier to interpret.

This step simply translate the gene IDs in the enzyme complexes into gene
symbols:

```
sub_network <- translate_complexes(sub_network)
```

Co-factors are removed. since we are aiming to follow metabolic routes
in the metabolic reaction network, it is crucial to remove over-promiscuous
metabolites that create unnecessary complexity in the network.

```
sub_network_nocofact <- remove_cofactors(sub_network)
```

Filter out any gene that isn't expressed from the metabolic network:

```
non_expressed_genes <- expressed_genes[rowSums(expressed_genes[,c(2),drop = F]) == 0,c(1),drop = F]
non_expressed_genes <- c(non_expressed_genes$gene)

tokeep <- list()
for(i in 1:length(sub_network_nocofact$reaction_network[,1]))
{
  elements <- gsub("[><].*","",sub_network_nocofact$reaction_network[i,])
  elements <- elements[!grepl("cpd:",elements)]
  elements <- unlist(strsplit(elements, "_"))
  tokeep[[i]] <- sum(elements %in% non_expressed_genes) == 0
}
tokeep <- unlist(tokeep)
sub_network_nocofact$reaction_network <- sub_network_nocofact$reaction_network[tokeep,]
```

Compress transporters to simplify the network:

```
sub_network_nocofact <- compress_transporters(sub_network_nocofact = sub_network_nocofact)
```

Split transaminases to preserve correct metabolic transformation routes:

```
sub_network_nocofact <- split_transaminases(sub_network_nocofact = sub_network_nocofact)

enzymes <- unique(sub_network_nocofact$attributes$V1)
enzymes <- enzymes[!grepl("_[clxmenr]$",enzymes)]
```

Convert the network into a forest (list of enzymes (trees) with correspoding
metabolic signatures (branches)):

```
sub_forest <- forestMaker(enzymes, sub_network_nocofact$reaction_network, branch_length = c(1,1), remove_reverse = T)
```

## Prepare the metabolic enzyme sets

```
penalty_min <- 6 #minimum 1 and integer
penalty_max <- 8 #maximum 9 and integer

reaction_set_list <- prepare_metabolite_set(
  penalty_range = penalty_min:penalty_max,
  forest = sub_forest,
  measured_metabolites = t_table$KEGG
)

reaction_set_list_merged <- condense_metabolite_set(reaction_set_list = reaction_set_list)

penalty <- 8 #has be between penalty_min and penalty_max and integer

regulons_df <- prepare_regulon_df(
  reaction_set_list_merged,
  penalty,
  filter_imbalance = c(0,1)
)
```

## Compute metabolic enzme enrichment score

```
metactivity_res <- metactivity(metabolomic_t_table = t_table,
                               regulons_df = regulons_df,
                               compartment_pattern = "_[a-z]$",
                               k = 1000)

mean_ES_df <- metactivity_res$ES

mean_NES_df <- metactivity_res$NES
```

```
translated_results <- translate_results(regulons_df = regulons_df, t_table = t_table, mapping_table = mapping_table, compress_compartments = T)
```

## Visualise results for single enzymes

```
plots <- plotMetaboliteContribution(
  enzyme = 'OGDH_DLD_PDH_DLST>1052',
  stat_df = translated_results$t_table,
  metabolite_sets = translated_results$regulons_df,
  contrast_index = 1,
  stat_name = 'Abundance Down <==> Up (t-value)',
  scaling_factor = 6,
  nLabels =  25
)

plot(plots$scatter)

plots <- plotMetaboliteContribution(
  enzyme = 'BCAT1',
  stat_df = translated_results$t_table,
  metabolite_sets = translated_results$regulons_df,
  contrast_index = 1,
  stat_name = 'Abundance Down <==> Up (t-value)',
  scaling_factor = 6,
  nLabels =  25
)

plot(plots$scatter)
```

```{r, warning=FALSE, message=FALSE}
sessionInfo()
```
